Classless Java

### 2015-11-06 paper/paper.tex
revised overview, with subsection structure
type checking of interfaces
Comments.
Going forward: formalization

% "define" code highlights for Java and Scala
\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,with,yield},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

@@

%{\theorembodyfont{\sffamily} \newtheorem{theorem}{Theorem}}
%{\theorembodyfont{\sffamily} \newtheorem{lemma}{Lemma}}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newenvironment{proof}{\textbf{Proof:\hspace{4mm}}}{$\Box$}
\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{pink}{#1}}
\newcommand{\hl}[1]{\textcolor{red}{#1}}
\newcommand\marco[1]{\authornote{marco}{blue}{#1}}

@@

\newcommand\mixin{\textbf{@Mixin}}
\usepackage{xspace}

>>>>

% "define" code highlights for Java and Scala
\lstdefinelanguage{JavaScala}{
  morekeywords={public,int,interface,implements,default,
    abstract,case,catch,class,def,static,%
    do,else,extends,false,final,finally,%
    for,if,implicit,import,match,mixin,%
    new,null,object,override,package,%
    private,protected,requires,return,sealed,%
    super,this,throw,trait,true,try,%
    type,var,while,yield},
  otherkeywords={=>,<-,<\%,<:,>:,\#,@},
  sensitive=true,
  morecomment=[l]{//},
  morecomment=[n]{/*}{*/},
  morestring=[b]",
  morestring=[b]',
  morestring=[b]"""
}

@@

%{\theorembodyfont{\sffamily} \newtheorem{theorem}{Theorem}}
%{\theorembodyfont{\sffamily} \newtheorem{lemma}{Lemma}}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}{Lemma}
%\newenvironment{proof}{\textbf{Proof:\hspace{4mm}}}{$\Box$}
\newcommand{\authornote}[3]{{\color{#2} {\sc #1}: #3}}
\newcommand\bruno[1]{\authornote{bruno}{red}{#1}}
\newcommand\yanlin[1]{\authornote{yanlin}{purple}{#1}}
\newcommand{\hl}[1]{\textcolor{red}{#1}}
\newcommand\marco[1]{\authornote{marco}{blue}{#1}}
\newcommand\haoyuan[1]{\authornote{haoyuan}{cyan}{#1}}

@@

\newcommand\mixin{\textbf{@Mixin} }
\usepackage{xspace}

### 2015-11-06 paper/sections/Formalization.tex
Going forward: formalization

\\[5ex]
\inferrule[(T-Obj)]{
\forall i\in 1..n\
%\Gamma_i
\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
\vdash\e_i\in \_\subtype\C^{\mh_i}\\\\
\forall i\in 1..n\ \mh_i\subtype\mBody(\m^{\mh_i},\C)\\\\
\forall i\in 1..k\ \Gamma(\x_i)\subtype\T_i\\\\
\forall\m\mbox{ such that }
\mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m
%\forall i\in 1\ldots n\ \Gamma_i=\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
}{
\Gamma \vdash\QM{new}\ \C\oR\cR\oC\T_1\ \f_1\QM=\x_1\QM;\ldots\T_k\ \f_k\QM=\x_k\QM;\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
\in\C
}
\quad
\inferrule[(T-Var)]{
\Gamma(\x)=\C
}{
\Gamma \vdash x \in\C}
\\[5ex]
 \inferrule[(T-Intf)]{
IT(\C_0) = \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \C_1\ldots\C_n \oC\methods\cC\\\\
 \forall \QM{default}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \Gamma^{\mh},\,\QM{this}{:}\C_0\vdash\e\in \_\subtype\C^{\mh} \\\\
 \forall \QM{static}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \Gamma^{\mh}\vdash\e\in \_\subtype\C^{\mh} \\\\
\dom(\C_0)=\dom(\C_1)\cup\ldots\cup\dom(C_n)\cup\dom(\methods)
 }{
\C_0 \text{ OK}
}
\end{array}$
\caption{Typing}
\label{ET}
\end{figure}
\subsection{Grammar and typing rules}${}_{}$\\

@@

We formalize a minimal version of Java 8, focusing on Interfaces, default methods and object creation literals.
As you can see, we have no syntax for classes.
Expressions are the conventional variable and method call$\x$ and$\e\QM.\m\QM(\es)$, then we have conventional static method call
$\C\QM.\m\QM(\es)$; for simplicity we do not consider the degenerate case of calling a static method over the this receiver as it considered a bad/confusing programming practice.
Next we have the more interesting super call $\C\QM{.super.}\m\QM(\es)$, whose semantic is to call the (non static) method $\m$ over the $\this$ receiver, but statically dispatching to the version of the method defined in the interface $\C$.
Finally, we consider the object initialization expression from an interface $\C$, where (for simplicity) all the fields are initialized with a variable present in scope.
Note how our language is exactly a subset of Java 8.
We allows the only annotation ... \marco{later}
We consider an globally present Interface Table (\metaVar{IT}) mapping from interface names $\C$ to interface declarations \metaVar{I}.

@@

We discuss first the most interesting rules, that is
(t-Obj) and  (t-Intf) \marco{to complete}
\marco{if we have space we can discuss the other rules too, but is not mandatory}

@@

\subsection{Translation Function}

$[\![\mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs \oC \methods \cC ]\!] =$\\*
${}_{}$\quad$\emptyset\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC 
\methods\ 
\ofMethod(\C_0) \ 
\otherMethod(\C_0)\ \cC
$\\*\tab
where  $\valid(\C_0)$
%\item $\overline{T\ m\oR\cR\QM;} = \fieldsFunc(C_0)$
%\item $p_i = \isWith(\mh^{\mBody(\QM{with#}m_i, C_0)}, C_0)$
%\item $q = \isClone(\mh^{\mBody(\clone, C_0)}, C_0)$
%\end{itemize}

@@

\subsubsection{Auxiliary function: $\valid, \ofMethod$, $\otherMethod$}${}_{}$\\*

\noindent$\valid(C_0)$  holds if $\forall \m\in\dom(\C_0),$ if $\mh\QM; = \mBody(\m, \C_0),$ one case is satisfied:
$\isField(\method)$,
$\isWith(\method, \C_0)$
or $\isClone(\method, \C_0)$.
%or $\isImplemented(\method)$.

We write $\QM{with#}\m$ to append $\m$ to \QM{with}, following the camelCase rule, so the first letter of
$\m$ must be lowercase and is turned in uppercase upon merging.
For example \QM{with#foo}=\QM{withFoo}.
Special names $\specialName(\m)$ are \QM{clone},\QM{with} and all the identifiers of form $\QM{with#}\m$.

We formally define $\ofMethod$, the function that generates the method \QM{of}, that behaves like a factory. To avoid boring digressions about well known ways to find unique names, for the sake of this formalization we assume that all the methods do not start with underscore, and we prefix method names with underscore to obtain valid field/parameter names.\\*
\noindent$\begin{array}{l}
\ofMethod(\C_0) = \
 \QM{static}\ C_0\ \QM{of} \oR \C_1\ \m_1\QM,\ldots \C_n\ \m_n\cR\
\QM{\{}
\QM{return new}\ C_0 \oR\cR\ \QM{\{} \\
\tab\tab\tab\tab\tab\tab\tab C_1\ \QM_\m_1 = \m_1\QM;\ldots \C_n\ \QM_\m_n = \m_n\QM; \\
\tab\tab\tab\tab\tab\tab\tab
\C_1\ \m_1\oR\cR\ \QM{\{return _}\m_1\QM{;\}}\ \ldots 
\C_n\ \m_n\oR\cR\ \QM{\{return _}\m_n\QM{;\}}\\
\tab\tab\tab\tab\tab\tab\tab\withMethod(\m_1,\C_0,\es_1)\ldots\withMethod(\m_n,\C_0,\es_n)\\
\tab\tab\tab\tab\tab\tab\tab\cloneMethod(\C_0,\es)\\
\tab\tab\tab\tab\tab\tab\QM{\};\}} \\
\end{array}$
\\*
with $\fieldsFunc(\C_0)=\C_1\ \m_1\QM{();},\ldots \C_n\ \m_n\QM{();}$,\\*
$\es_i=\QM{this.}\m_1\QM{(),}\ldots \QM{this.}\m_{1-i}\QM{(),}\QM{_val,}\QM{this.}\m_{i+1}\QM{(),}\ldots \QM{this.}\m_n\oR\cR$\\*
and $\es=\QM{this.}\m_1\QM{(),}\ldots\ldots \QM{this.}\m_n\oR\cR$\\*

\noindent{}Other auxiliary functions:\\*
\noindent$\begin{array}{ll}
\withMethod:&\withMethod(\C,\m,\C_0,\es)=\method=
\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\ \QM{\{}
\QM{return} \C_0\QM{.of(}\es\QM{);\}} \\
&\mbox{iff }\method<:\mBody(\QM{with#}\m,\C_0)\\
&\withMethod(\C,\m,\C_0,\es)=\emptyset\mbox{ otherwise}\\
\cloneMethod:&\cloneMethod(\C_0,\es)=\method=
\C_0\ \QM{clone()\{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
&\mbox{iff }\method<:\mBody(\QM{clone},\C_0)\\
&\cloneMethod(\C_0,\es)=\emptyset\mbox{ otherwise}\\

\otherMethod:& \C_0\ \QM{with#}\m\oR \C_i\ \x_i\cR\QM;\in
\otherMethod(\C_0) 
\\&
 \mbox{iff }
\C\ \m\QM{();}\in \fieldsFunc(\C_0)\mbox{ and } \QM{with#}\m\notin\dom(\C_0)\\
&\C_0\ \QM{clone}\oR\cR\QM;\in
\otherMethod(\C_0)  \mbox{ iff } \isClone( \mBody(\m, \C_0))\\

%other methos was wrong: we need the with to be NOT there to generate it.
%no use isWith, any method with that name must be absent

\fieldsFunc:&\method\in\fieldsFunc(\C_0) \mbox{ iff }
\method\in \dom(\C_0)\ \mand\ \isField(\method)
\\
\isField:&\isField(\C\ \m\oR\cR\QM;)\tab \mif\ \mnot\ \specialName(\m)\\
\isWith:&\isWith(\C'\ \QM{with#}\m \oR \C\ \x\cR\QM;, \C_0)
\tab \mif\ \C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;\\
& \mand\ \mnot\ \specialName(\m)\\
\isClone:&\isClone(\C\ \QM{clone}\oR\cR\QM;, \C_0)\tab \mif\ \C_0 <: \C \\
%\isImplemented:&\isImplemented(\method) \tab\mbox{iff }\method\mbox{ not of form }\mh\QM;
%\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}) = \QM{true} \\
%&\isImplemented(\QM{static}\ \mh\mbox{\Q@\{return \_;\}@}) = \QM{true} \\
\end{array}$

\subsubsection{Results}
\textbf{THEOREM. } 
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK, then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n, \II$ OK.

\subsubsection{Derived notations}

 Below shows how the functions $mtype$ and $mmodifier$ are derived from $mbody$.

\[ \textsf{mbody}(m,C) = \textit{modifier } E \spc m(\overline{D} \spc \overline{x}) \{ \text{return } e; \} \] \[ \Rightarrow \textsf{mtype}(m,C) = \overline{D} \to E,\ \textsf{mmodifier}(m,C) = \textit{modifier}\]

>>>>

\\[5ex]
\inferrule[(T-Obj)]{
\forall i\in 1..n\
%\Gamma_i
\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
\vdash\e_i\in \_\subtype\C^{\mh_i}\\\\
\forall i\in 1..n\ \mh_i\subtype\mBody(\m^{\mh_i},\C)\\\\
\forall i\in 1..k\ \Gamma(\x_i)\subtype\T_i\\\\
\forall\m\mbox{ such that }
\mBody(\m,\C)=\mh\QM; \exists i\in 1..n\ \m^{\mh_i}=\m
%\forall i\in 1\ldots n\ \Gamma_i=\Gamma,\f_1{:}\T_1,\ldots,\f_k{:}\T_k,\,\QM{this}{:}\C,\Gamma^{\mh_i}
}{
\Gamma \vdash\QM{new}\ \C\oR\cR\oC\T_1\ \f_1\QM=\x_1\QM;\ldots\T_k\ \f_k\QM=\x_k\QM;\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
\in\C
}
\quad
\inferrule[(T-Var)]{
\Gamma(\x)=\C
}{
\Gamma \vdash x \in\C}
\\[5ex]
 \inferrule[(T-Intf)]{
IT(\C_0) = \ann\ \QM{interface}\ \C_0\ \QM{extends}\ \C_1\ldots\C_n \oC\methods\cC\\\\
 \forall \QM{default}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \Gamma^{\mh},\,\QM{this}{:}\C_0\vdash\e\in \_\subtype\C^{\mh} \\\\
 \forall \QM{static}\ \mh\oC\QM{return}\ \e\QM;\cC \in \methods,
\ \Gamma^{\mh}\vdash\e\in \_\subtype\C^{\mh} \\\\
\dom(\C_0)=\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)
 }{
\C_0 \text{ OK}
}
\end{array}$
\caption{Typing}
\label{ET}
\end{figure}
\subsection{Grammar and typing rules}${}_{}$\\

@@

We formalize a minimal version of Java 8, focusing on Interfaces, default methods and object creation literals.
As you can see, we have no syntax for classes.
Expressions are the conventional variable and method call$\x$ and$\e\QM.\m\QM(\es)$, then we have conventional static method call
$\C\QM.\m\QM(\es)$; for simplicity we do not consider the degenerate case of calling a static method over the this receiver as it considered a bad/confusing programming practice.
Next we have the more interesting super call $\C\QM{.super.}\m\QM(\es)$, whose semantic is to call the (non static) method $\m$ over the $\this$ receiver, but statically dispatching to the version of the method defined in the interface $\C$.
Finally, we consider the object initialization expression from an interface $\C$, where (for simplicity) all the fields are initialized with a variable present in scope.
Note how our language is exactly a subset of Java 8.
We allows the only annotation \mixin since is the only one interesting in this article.
We consider an globally present Interface Table (\metaVar{IT}) mapping from interface names $\C$ to interface declarations \metaVar{I}.

@@

We discuss first the most interesting rules, that is
\rn{(t-Obj)} and  \rn{(t-Intf)} \marco{Bruno, you can explain t-obj, if you can make it understandable, I go for t-inft}

Rule \rn{(t-Obj)}....

Rule \rn{(t-intf)} check that an interface $\C_0$ is correctly typed.
First we check that the body of all the default and static methods are well typed.
Then we check that
$\dom(\C_0)$ is the same of $\dom(\C_1)\cup\ldots\cup\dom(\C_n)\cup\dom(\methods)$.
This is not a trivial check, since $\dom(\C)$ is defined using $\mBody$, that is undefined in many cases: notably if a method $\method\in\methods$ is not compatible with some method in
$\dom(\C_1)\ldots\dom(\C_n)$ or if any method in both $\dom(\C_i)$ and $\dom(\C_j)$ ($i,j\in 1..n$)  is conflicting.

\marco{if we have space we can discuss the other rules too, but is not mandatory}

@@

\subsection{What  \mixin Generates}
We now show what the \mixin annotation generates. We present a formal definition for
most of the generated methods; however in our formalism we do not consider setters, so we do not formally define the generation of the setters. 
We do not include casts or \Q@instanceof@, so we do not include the \Q@with@ method.

\subsubsection{Translation Function}${}_{}$\\*

\noindent$[\![\mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs \oC \methods \cC ]\!] =
\emptyset\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC 
\methods\ \methods' \cC
$\\*${}_{}$\tab
where  $\valid(\C_0)$,\Q@of@$\notin\dom(\methods)$ and $\methods'=\ofMethod(\C_0) \ \otherMethod(\C_0,\methods)$.\\*

To translate an annotated interface, we add the \Q@of@ method, and then we add some other methods.
However, first of all we check if the interface is valid for annotation:

\noindent$\valid(\C_0)$  holds if $\forall \m\in\dom(\C_0),$ if $\mh\QM; = \mBody(\m, \C_0),$ one case is satisfied:
$\isField(\method)$,
$\isWith(\method, \C_0)$
or $\isClone(\method, \C_0)$.

That is, we can categorize all the \emph{not implemented} methods in a pattern that we know how to implement.
To complete our formal definition we would need to add setters and \Q@with@.
Moreover, we check that the method \Q@of@ is not already defined by the user.
In our simplified formalization we consider this to be just an error.
In our prototype we keep \Q@overloading@ into account, and so we check that an of method with the same signature of the one we would like to generate is not already present.\marco{Do we check it?}

In the following we will write $\QM{with#}\m$ to append $\m$ to \QM{with}, following the camelCase rule, so the first letter of
$\m$ must be lower-case and is turned in upper-case upon merging.
For example \QM{with#foo}=\QM{withFoo}.
Special names $\specialName(\m)$ are \QM{clone}, \QM{with} and all the identifiers of form $\QM{with#}\m$.

\subsubsection{$\ofMethod$}${}_{}$\\*
We now formally define $\ofMethod$, the function that generates the method \QM{of}, that behaves like a factory. To avoid boring digressions about well known ways to find unique names, for the sake of this formalization we assume that all the methods do not start with underscore, and we prefix method names with underscore to obtain valid  parameter names.\\*
\noindent$\begin{array}{l}
\ofMethod(\C_0) = \
 \QM{static}\ C_0\ \QM{of} \oR \C_1\ \QM_\m_1\QM,\ldots \C_n\ \QM_\m_n\cR\
\QM{\{}
\QM{return new}\ C_0 \oR\cR\ \QM{\{} \\
\tab\tab\tab\tab\tab\tab\tab C_1\ \m_1 = \QM_\m_1\QM;\ldots \C_n\ \m_n = \QM_\m_n\QM; \\
\tab\tab\tab\tab\tab\tab\tab
\C_1\ \m_1\oR\cR\ \QM{\{return }\ \m_1\QM{;\}}\ \ldots 
\C_n\ \m_n\oR\cR\ \QM{\{return }\ \m_n\QM{;\}}\\
\tab\tab\tab\tab\tab\tab\tab\withMethod(\m_1,\C_0,\es_1)\ldots\withMethod(\m_n,\C_0,\es_n)\\
\tab\tab\tab\tab\tab\tab\tab\cloneMethod(\C_0,\es)\\
\tab\tab\tab\tab\tab\tab\QM{\};\}} \\
\end{array}$
\\*
with $\fieldsFunc(\C_0)=\C_1\ \m_1\QM{();},\ldots \C_n\ \m_n\QM{();}$,\\*
$\es_i=\m_1\QM,\ldots\QM, \m_{i-1}\QM,\QM{_val,}\m_{i+1}\QM,\ldots\QM, \m_n$\\*
and $\es=\m_1\QM,\ldots\QM, \m_n$\\*

The function $\fieldsFunc(\C_0)$ (formally defined later) denotes all the fields in the current interface.
For methods inside the interface with the form $\C_i\ \m_i$\QM{();}
  \begin{itemize}
   \item $\m_i$ is the field name, and have type $\C_i$.
   \item $\m_i$\QM{()} is the getter, that just return the current field value.
   \item if a method \Q@with#@$\m_i$ is required, then it is implemented by calling the \Q@of@ method using
    the current value for all the fields except for $\m_i$. Such new value is provided as parameter. This correspond to the expressions $\es_i$.
   \item similarly, for the \Q@clone@ method, \Q@of@ is called using the current value for all the fields.
   \item To complete our generation, we need to generate setters, fluent setters and the with method.
   \item \marco{should we just formalize setters?}
   \end{itemize}

\subsubsection{Other auxiliary functions}${}_{}$\\*

\noindent$\begin{array}{ll}
\withMethod:&\withMethod(\C,\m,\C_0,\es)=
\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\ \QM{\{}
\QM{return} \C_0\QM{.of(}\es\QM{);\}} \\
&\mbox{iff }
\mBody(\QM{with#}\m,\C_0) \mbox{ is of form }\mh\QM;\\
&\withMethod(\C,\m,\C_0,\es)=\emptyset\mbox{ otherwise}\\
\cloneMethod:&\cloneMethod(\C_0,\es)=
\C_0\ \QM{clone()\{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
&\mbox{iff }
\mBody(\QM{clone},\C_0) \mbox{ is of form }\mh\QM;\\
&\cloneMethod(\C_0,\es)=\emptyset\mbox{ otherwise}\\
\end{array}$
As you can see above, \Q@with-@ and \Q@clone@ methods are generated if needed.
We can discover if there is the need of generating such methods by checking if the method is unimplemented in $\C_0$. Note that we do not need to check if its header is a subtype of what we would generate, this is ensured by $\valid(\C_0)$.

\noindent$\begin{array}{ll}
\otherMethod:& \C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods) 
\\&
 \mbox{iff }
\C\ \m\QM{();}\in \fieldsFunc(\C_0), \isWith(\mBody(\QM{with#}\m, \C_0)) 
\\&\mbox{ and } \QM{with#}\m\notin\dom(\methods)\\
&\C_0\ \QM{clone}\oR\cR\QM;\in
\otherMethod(\C_0)
\\&  \mbox{ iff } 
\isClone(\mBody(\QM{clone}, \C_0)) 
\mbox{ and } \QM{clone}\notin\dom(\methods)\\
\end{array}$

Other methods that we need to generate in the interface are \Q@with-@ and \Q@clone@.
A complete formalization would also generate the \Q@with@.
This is needed only if we need to refine the return type.
To discover if this is the case, we check if such \Q@with-@ or \Q@clone@ is required by $\C_0$, but is not already present in the methods directly declared in $\C_0$.

\noindent$\begin{array}{ll}
\fieldsFunc:&\method\in\fieldsFunc(\C_0) \mbox{ iff }
\method\in \dom(\C_0)\ \mand\ \isField(\method)
\\
\isField:&\isField(\C\ \m\oR\cR\QM;)\tab \mif\ \mnot\ \specialName(\m)\\
\isWith:&\isWith(\C'\ \QM{with#}\m \oR \C\ \x\cR\QM;, \C_0)
\tab \mif\ \C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;\\
& \mand\ \mnot\ \specialName(\m)\\
\isClone:&\isClone(\C\ \QM{clone}\oR\cR\QM;, \C_0)\tab \mif\ \C_0 <: \C \\
%\isImplemented:&\isImplemented(\method) \tab\mbox{iff }\method\mbox{ not of form }\mh\QM;
%\QM{default}\ \mh\mbox{\Q@\{return \_;\}@}) = \QM{true} \\
%&\isImplemented(\QM{static}\ \mh\mbox{\Q@\{return \_;\}@}) = \QM{true} \\
\end{array}$

Our prototype also generate setters; informally
\begin{itemize}
\item For methods inside the interface with the form \Q@void @$\m$\QM($\C\ \x$\QM{);}:
  \begin{itemize}
    \item Check if exist method $\C\ \m$\Q@();@. If not, generate error (that is, is not $\valid(\C_0)$).
    \item Generate implemented setter method inside \Q@of@:\\*
           \Q@public void @$\m$\Q@(@$\C$\Q@ _val) { @$\m$\Q@=_val;}@
    \end{itemize}
\item For methods with the form $\C'\ \m$\QM($\C\ \x$\QM{);}:
  \begin{itemize}
    \item As for before, check if exist method $\C\ \m$\Q@();@. Also, check that $\C'$ is a supertype of the current interface type $\C_0$.
    \item Generate implemented setter method inside \Q@of@:\\*
           \Q@public @$\C_0\ \m$\Q@(@$\C$\Q@ _val) { @$\m$\Q@=_val; return this;}@
    \item If needed, as for \Q@with-@ and clone, generate the method header with refined return type in the interface.
  \end{itemize}
\end{itemize}

Note how the second kind of setter is what is called a fluent setter~\cite{the lombock thread, something about fluent stuff}.
This allows for convenient and chains of setters, as we will show later \marco{insert forward reference when available}.

\subsubsection{Results}
\textbf{THEOREM. } 
For a given $\II_0\ldots\II_n$ interface table such that
$\forall\II\in\II_0\ldots\II_n, \II$ OK, then in the interface table
$[\![\II_0]\!]\II_1\ldots\II_n$
$\forall\II\in[\![\II_0]\!]\II_1\ldots\II_n$ either $\II$ OK or $\II$ is a subtype of $\II_0$.

To understand this theorem statement, we need to understand that there are three kind of guarantees that we can offer
\marco{should we repeat the 3 points in the email?}

%\subsubsection{Derived notations}
% Below shows how the functions $mtype$ and $mmodifier$ are derived from $mbody$.
%\[ \textsf{mbody}(m,C) = \textit{modifier } E \spc m(\overline{D} \spc \overline{x}) \{ \text{return } e; \} \] \[ \Rightarrow \textsf{mtype}(m,C) = \overline{D} \to E,\ \textsf{mmodifier}(m,C) = \textit{modifier}\]

@@

### 2015-11-06 paper/sections/Overview.tex
Going forward: formalization

Just to give you a feeling on how much boring code \mixin is generating,  we show the correspondent code without then \mixin annotation.\footnote{Can you spot our intentional typo? This kind of code is so boring that our brain refuse to process it. Programmers should not be required to write this kind of code.}
\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);
  Point3D withY(int val);
  Point3D withZ(int val);
  public static Point3D of(int x, int y, int z){return new Point3D(){
    public int x(){return x;} 
    public int y(){return y;}
    public int z(){return z;}
    public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
    public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
    public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
    }
  }
}
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.

@@

@@

Note how
\begin{itemize}
\item by covariant method overriding we refine the type of our field
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}
The cast is trivially safe since it is guarded by an \Q@instanceof@ test.
The chain of \Q@with-@ methods is not fully satisfactory: it requires \Q@Bird@ to know of all the fields of \Q@Point2D@. We are planning to extend our \mixin annotation so that 
we can write \Q@point(point().with(val))@, where the \Q@with(T val)@ method would behave like calling all the \Q@with-@ where a compliant getter exists in \Q@T@.
\marco{We need to discuss this, we may just implement it? --in this case the text would just become:}
In Point3D:
\begin{lstlisting}
@Mixin interface Point3D extends Point2D{
  int z(); Point3D withZ(int val);
  Point3D with(Point val);
  }
\end{lstlisting}
expandend in 
\begin{lstlisting}
interface Point3D extends Point2D{
  ....
  Point3D with(Point2D val){
    if(val instanceof Point3D){return (Point3D)val;}
    return Point3D.of(val.x(),val.y(),this.z());
    }
  ...
  }
\end{lstlisting}
and here some discussion about the new with method,
and then the Bird becomes 
\begin{lstlisting}
@Mixin interface Bird extends Animal{
  Point3D point();
  void point(Point3D val);
  default void point(Point2D val){ return point(point().with(val));}
  default void fly(){...}
  }
\end{lstlisting}
\marco{what do we think?}
\subsubsection{\Q@Pegasus@ and multiple Inheritance}
The \emph{``multiple inheritance''} case appears at interface
\texttt{Pegasus}.
Pegasus (one of the best known creatures
in Greek mythology) can not only \emph{run} but also \emph{fly}! 
It would be handy if they could obtain \texttt{fly} and \texttt{run} functionality from both \texttt{Horse} and \texttt{Bird}.

@@

\subsection{What  \mixin Generates}
Apart from generating code for \emph{getter} methods, the annotation processor
can also generate \emph{setter} methods, \emph{with} methods, \texttt{clone}
methods, etc. We summarize all the forms of method \mixin  supports below. Inside
the anonymous class in the annotated interface, the following code are
generated:

\begin{itemize}
\item For methods inside the interface with the form \texttt{Tx x()}:
  \begin{itemize}
   \item \texttt{x} is the getter method, with return type
     \texttt{Tx}. Conceptually, it is a member field with name \texttt{x} and
     type \texttt{Tx}.
   \item generate member field \texttt{\_x} of type \texttt{Tx}, initialized
     with \texttt{x}.
   \item generate implemented getter method:
       \begin{lstlisting}
           public Tx x() { return _x; }
       \end{lstlisting}
   \end{itemize}

\item For methods inside the interface with the form \texttt{void x(Tx x)}:
  \begin{itemize}
    \item check if exist method \texttt{Tx x()}. If not, generate error.
    \item generate implemented setter method:
       \begin{lstlisting}
           public void x(Tx x) { this.x = x; }
       \end{lstlisting}
    \end{itemize}

\item For methods with the form \texttt{T withX(Tx \_)}:
  \begin{itemize}
  \item if there is no \texttt{x} field, or type \texttt{Tx} does not match,
    then generate error.
  \item implement `withX` using the `of` method.
  \end{itemize}

\item For methods with the form of \texttt{T clone()}: Use \texttt{of} method as
  the constructor, to create a new object with the same field values as the
  current one.

\item For methods with the form of \texttt{T x(Tx \_)}:
  \begin{itemize}
    \item check if exist method \texttt{T x()}, if not, generate error.
    \item inside the inner class, generate
      \begin{lstlisting}
          public T x(Tx x) { this.x = x; return this;}
      \end{lstlisting}
  \end{itemize}
\end{itemize}

>>>>

Just to give you a feeling on how much boring code \mixin is generating,  we show the correspondent code without then \mixin annotation.\footnote{Can you spot our intentional typo? This kind of code is so boring that our brain refuse to process it. Programmers should not be required to write this kind of code.}
\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);
  Point3D withY(int val);
  Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int x, int y, int z){return new Point3D(){
    public int x(){return x;} 
    public int y(){return y;}
    public int z(){return z;}
    public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
    public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
    public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
    public Point3D with(Point2D val){
      if(val instanceof Point3D){return (Point3D)val;}
      return Point3D.of(val.x(),val.y(),this.z());
      }
    };
  }
}
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

@@

\lstinputlisting[linerange=80-88]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

@@

Note how
\begin{itemize}
\item by covariant method overriding we refine the type of our field
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}

From the type perspective, the key is the covariant method overriding; however from the semantic perspective the key is the implementation for the setter with the old signature:
by using the \Q@with@ method we may use the information for \Q@z@ already stored in the object to forge an appropriate \Q@Point3D@ to store.
Note how all the informations about what fields sits in \Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the \Q@with@ method, and is transparent for the implementer of \Q@Bird@.

\subsubsection{\Q@Pegasus@ and multiple Inheritance}
The \emph{``multiple inheritance''} case appears at interface
\texttt{Pegasus}.
Pegasus (one of the best known creatures
in Greek mythology) can not only \emph{run} but also \emph{fly}! 
It would be handy if they could obtain \texttt{fly} and \texttt{run} functionality from both \texttt{Horse} and \texttt{Bird}.

@@
